---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# HINF5300

<!-- badges: start -->
<!-- badges: end -->

This package documents any code and assignments for the HealthInformatics5300
course at Northeastern University.

## Installation

You can install the development version of HINF5300 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("TinasheMTapera/HINF5300")
```

## Assignment 1: Step Detection

This assignment was demonstrating how to detect steps from an accelerometer
embedded in a mobile device. The notebook for the assignment is located
at `vignette("assignment1")`:

```{r assignment1-example, warn=FALSE, message=FALSE}
library(HINF5300)
library(dplyr)
## basic example code

steps_df <- detect_steps(system.file("extdata", "acce.csv", package = "HINF5300"))
steps_df %>%
  summarise(n_steps = sum(!is.na(step)))
```
