---
title: "Try it Yourself!"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Try it Yourself!}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HINF5300)
```

## Installation

You can install this package from Github using `remotes`:

```{r, eval=FALSE}
# install.packages("remotes") # if necessary
remotes::install_github("TinasheMTapera/HINF5300")
```

Or with the source files, by opening this `.Rproj` file in RStudio and clicking
the `Install` button under the Build tab.

## Assignment 1

You can try assignment 1's step detection algorithm by first using the Datalogger
iOS or Android app to collect raw accelerometer data. Then, once that CSV file is on your machine with this package installed,
simply do `HINF5300::runExamples("assignment1", <DESIRED OUTPUT DIRECTORY>)`. A pop up window will ask you
to select your data and set the parameters for step detection as described in the
notebook. Submitting the data will produce a pretty step detection report for you. It's also possible to do this in R with `HINF5300::detect_steps(<YOUR FILE>, <PARAMS>)` which will return a data frame that you can plot yourself.
